Select-based random access \\ to variable-byte encodings

### 2020-05-24 gradu/msc_english_contents.tex
Final fixes

\chapter{Introduction}
Enormous datasets are a common case in today's applications. Some datasets are too large to handle in memory and thus must be stored in another, slower storage \citep{Hon10}. 
Compressing these datasets is often beneficial, because doing so naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95, Wil99}.

Fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
data compression algorithms are inefficient. Fortunately, such random access to elements is not a requirement compression algorithms usually 
need to fulfill.

However, efficient random access to compressed data is very useful in compressed data structures \citep{Ven13}. It saves storage space, bandwidth and 
increases the likelihood of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data 
from the beginning.

Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuation bit} is added to the data to denote whether the compressed integer
is continued to the next byte or not. Two different light-weight data structures are built over the continuation bit array. $Rank_1$($i$) returns the number of 1 bits in the bit array between 
indexes 0 and $i$-1. $Select_1$($i$) returns the index of the i-th 1 bit in the array. The fact that the compressed data blocks are of the same size is used
with the data structures to allow direct access to the compressed variable-byte data.

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17, Sha16}.

In this thesis, an alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. The underlying question is if a $select$ based method is viable for random 
access. The proposed method does not use data 
reorganizing, but rather capitalizes on the assumption that data is encoded the same way it is in the source. Because variable-byte encoding does not assign 
new codewords to data when encoding, data can be read straight from the memory and reassembling the integer from blocks is not needed.

Comparison of the proposed method and $rank$-based implementation by Brisaboa et al is provided with different implementations of $rank$ and $select$. The $rank$ method 
is expected to perform better on data sets containing mostly small integers. It is assumed that larger numbers cause problems in performance. The performance of the proposed $select$ method is indifferent 
of the size of the elements in the data. Different kinds of data sets are used to see how each algorithm performs. The assumption is that $select$ works better with data sets with very large integers. It
remains to be seen how large the integers need to be for the $select$ method to outperform the $rank$ method.

Another interesting point is to compare the algorithms when decompressing a long subarray. Due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element 
and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.

@@

Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for the integer data sets. Their encoding process assigns short bit sequences to small integers, 
which is why they outperform in terms of compression VB encoding on datasets containing a lot of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. VB is generally faster 
to decode \citep{Wil99, Tro03} and so is sometimes preferred, also due to its simplicity of implementation. In this thesis, the data sets used for experimentation will have the focus on slightly larger, 
unordered integers.

@@

VB encoding was first experimented as a tool for compressing inverted index lists of word locations in documents by \citep{Sch02}. 
In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed to an inverted index, storing each number as its difference to the previous number 
instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, but inverted index encoding makes the integers are small. 
In contrast, rare words have only a few entries but the integers stored are larger. These lists are mostly small integers with a need to support larger ones and thus are excellent candidates for VB encoding.

@@

\section{VB encoding}
The name \textit{variable-byte} is reflected in how the integers are stored. When stored in an array, each element requires the same amount of space. This is very inefficient if a lot of the values are 
very small but the array also has to support very large integers. VB encoding attempts to eliminate the unneeded leading zeros. The bits in an integer are split into blocks of length $b$ and empty blocks 
from the beginning are discarded. Because the lengths of the integers may now be different, the data cannot be stored as it is. Instead a \textit{continuation bit} is added in front of each block to form 
a $chunk$. This bit is set to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer 
continues in the next chunk.

For example, the standard 16-bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. Assuming the block length is 4, it is split to \texttt{0000 0000 0010 1010}. The empty blocks 
are removed from the beginning and then the continuation bits are added. 1 is added in front of the last block (containing the least significant bits) and 0 to the other blocks, resulting in 
\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with byte length 4.

@@

A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length. Since the block 
length is 7, the continuation bit is added in row 8 by adding $2^8$ = 128 to the last block.

@@

Smaller block lengths can yield a better compression rate at the cost of more bit manipulation and therefore possibly slower decompression, while bigger block lenghts need less bit manipulation and 
offer less effective compression. On the other hand, bigger block length means a smaller percentage of added continuation bits.
Generally, a block length of 7 has been popular because it tends to perform well on average and handling chunks as bytes is 
convenient \citep{Man08}.

\section{VB decoding}
VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the resulting blocks are concatenated to form the original integer. A pseudo code implementation of VB decoding with a block length of 7 is shown in Figure~\ref{vbyte_dec}.
\textsc{Append} adds an element to the end of the list. If the block length is changed, additional bit operation steps when reading the data may be needed.

As how the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
extracted and added to $n$, making $n$ = \texttt{10} (bit representation of 2).
The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
changing $n = $\texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
$n = $\texttt{101010} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.

@@

To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect to the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
created in addition to the existing data. The length of the bit array in these cases usually has to be close to the length of the data, because encoded element lengths vary. This spatial increment makes 
the compression ineffective and therefore $rank$ and $select$ cannot be applied to any given compression algorithm.

VB encoding has several advantages with $search$ and $rank$: the data is compressed in 
blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
 of both rank and select to support the bit array. Also a few useful functions from the library were used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
size requirements of implementations used in this experiment. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
compressed sequence by much less than one bit per element. Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$. 
These implementations were chosen because they are fast and can be created over a regular bit array.

@@

For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, both division operations being integer divisions. Then number of 1 bits 
from index $r$ to $i$ are calculated from the original bit array. All three values added 
together add up to $rank$($i$). $Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, 
but reduces memory requirement to a quarter.

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits. 
The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.

The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each. The first relative block again equals to 0, so the rest 
of the blocks fit to a 64-bit word. The $rank_{v5}$ implementation requires an additional (64+64)/2048 = 6.25\% space.

The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
relative to the superblock. With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached. The $select_{MCL}$
 is a practical variant of a PAT tree \citep{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.

In both cases, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}. The 
implementations are explained in greater detail in \citep{gbmp2014sea}.

@@

The ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element in a compressed list or array is one of the usual requirements in compressed data 
structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.

It is not natively possible to decode the $i$-th element in variable-byte compression algorithms, because the position of the element in the compressed
list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution is to store the location 
of each element to an array, but it adds a very large overhead which removes the benefit of compression.

@@

After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

\end{tikzpicture}
\caption{Data structure by Brisaboa et al., visualized}\label{bris_ds}

@@

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ is the index of desired element's next block in array $A_2$. This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current 
block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.

As an example, the fourth element from the data structure shown in Figure~\ref{bris_ds} is decoded. The first block of data is fetched by directly accessing $A_1$[3]. Then the 
continuation bit $B_1$[3] is checked. It is not set, which indicates that the element continues to the next block. The index of the next block is obtained via a \textsc{Rank}($B_1$, 3) call. There are two zero 
bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is set to 1 so the decompression is completed.

Interestingly the block length of this method can change within different array levels which opens a possibility for further compression optimization. \citep{Bri09} experiments a few different block 
length setups and leaves open the problem of finding optimal block lengths. While the problem of finding the optimal block length for each data set in regular VB encoding is interesting, the possibility to
optimize it for each block level separately is surely more challenging.
\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $index \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}

@@

\chapter{DAC with select query}
\label{chap:select_impl}
Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

@@

Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own array, which leaves the blocks in their own array. This allows the 
compressed number to be read from the memory block and removes the need of block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
when reading a word from the block byte array, the bits and bytes are already in correct order.

For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$(i) gets the index location
of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
The first block might not be at the start of the byte $s$. This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.

At this point, $w$ contains the wanted bits, but also has extra bits from the previous and next compressed integers. If the offset was not zero, there are $o$ trailing bits in $w$ from the previous integer. 
These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the desire value. Most systems are
able to read memory only from byte addresses, which leads to a corner case where the compressed integer is stored in more bytes than are contained in a word. In the aforementioned situation, this 
happens if compressed length was 8 blocks and offset was 4. Then another step of reading another byte and storing the remaining bits to $w$ is required. This is entirely avoidable if the integer 
length is constrained. In this case the maximum integer length of 28 bits would eliminate the handling of the second byte.

The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
$select_1$ query, which is costly. A much faster option is to calculate the block length from the continuation bit array. This can be done by reading a word from the bit array and bit-shifting the word 
until the bit representing the start of the i-th element is on the rightmost end and then counting the trailing zeros. This counts the number of blocks in the i-th element that do not have the continuation bit set. 
This number is then incremented by one to also count the ending block which has the continuation bit set to 1.

Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the pseudo code, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns 
a bit mask for k bytes. In line 8, a long word is read from the memory starting at memory address $A$[$begin$].

@@

Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. The length of the compressed 
element is 3, which is obtainable, e.g., via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$ and blocks not belonging to the element
are removed either with a bitmask or by double bit shifting, resulting in k-th element being decoded.

@@

\draw[arrow] (Btarget.south) to (Atarget.north);
\end{tikzpicture}
\caption{Relation of continuation bits and data blocks in VB encoding} \label{VBarrays}
\end{figure}

>>>>

\chapter{Introduction}
Enormous datasets are a common case in today's applications. Some datasets are too big to handle in memory and thus must be stored in larger, slower storage \citep{Hon10}. 
Compressing these datasets is often beneficial, because doing so naturally decreases memory requirements, but also can make it faster to transfer data from disk to memory \citep{Zob95, Wil99}.

A fundamental method for data compression is variable-length coding \citep{Sal99}. The main idea of variable-length encoding is that 
frequent sequences of data are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element within the compressed data. If this is required, the usual 
data compression algorithms are inefficient, because data must be decompressed from the beginning to the point where the the sought element is located. 
Fortunately, such random access to elements is not a requirement compression algorithms usually need to fulfill.

However, efficient random access to compressed data is very useful in compressed data structures \citep{Ven13}.  In addition to saving storage space and bandwidth, the compression 
increases the likelihood of data already being in cache \citep{Sch02}.

Variable-byte encoding is a method for compressing integers of different byte length. An additional \textit{continuation bit} is added to the data to denote whether the compressed integer
is continued to the next byte or not. To facilitate random access, two different light-weight data structures are built over the continuation bit array. $Rank_1$($i$) returns the number of 1 bits in the bit array between 
indexes 0 and $i$-1. $Select_1$($i$) returns the index of the i-th 1 bit in the array. The fact that the compressed data blocks are of the same size is used
with the data structures to allow direct access to the compressed variable-byte data.

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[see, e.g.,][]{Kon17, Sha16}.

In this thesis, a novel alternative solution that instead makes use of a $select$ data structure is proposed and explained in detail. The underlying question is if a $select$ based method is viable for random 
access. Compared to \citeauthor{Bri09}, proposed method uses a very simple data 
reorganizing, and capitalizes on the assumption that data is encoded the same way it is in the source. Data can then be read straight from the memory and reassembling the integer from 
its encoded blocks is not needed.

Comparison of the proposed method and the $rank$-based implementation by \citeauthor{Bri09} is provided with different implementations of $rank$ and $select$. The $rank$ method 
is expected to perform better on data sets containing mostly small integers. It is assumed that larger numbers lead to slower random access for $rank$-based methods. 
The performance of the proposed $select$ method is indifferent 
of the size of the elements in the data. Different kinds of data sets are used to assess how each approach performs. The assumption is that $select$ works better with data sets with very large integers. It
remains to be seen how large the integers need to be for the $select$ method to outperform the $rank$ method.

The performance of the algorithms when decompressing a long subarray is also compared. Because of the way the data is stored, the proposed $select$ method offers fast access to the next or previous element 
and is expected to perform well with any given data set or subarray length. Solutions for subarray fetching with both $rank$ and $select$ methods are proposed and compared.

@@

Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for integer data sets. Their encoding process assigns short bit sequences to small integers, 
which is why they outperform, in terms of compression, VB encoding on datasets containing lots of small integers \citep{Wil99}, such as inverted indexes \citep{Anh05, Pib19}. VB is generally faster 
to decode \citep{Wil99, Tro03} and so is sometimes preferred, also due to its simplicity of implementation. In this thesis, the data sets used for experimentation will have the focus on slightly larger, 
unordered integers.

@@

VB encoding was first experimented as a tool for compressing inverted index lists of word locations in documents by \citeauthor{Sch02}. 
In that setting, VB codes yielded excellent results, and since then many different approaches have been introduced. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in the document for 
advanced search purposes. Usually these lists are preprocessed to form an inverted index, storing each number as its difference to the previous number 
instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, which makes the inverted index encoded integers small.
In contrast, rare words have only a few entries but the integers stored are larger. These lists are mostly small integers with a need to support larger ones and thus are excellent candidates for VB encoding.

@@

\section{Variable-byte encoding}
The name \textit{variable-byte} reflects the way in which the integers are stored. When the integers are stored in an array, each element requires the same fixed amount of space. This is inefficient if a lot of the values are 
very small but the array also has to support very large integers. VB encoding attempts to eliminate the unneeded leading zeros. The bits in an integer are split into blocks of length $b$ and empty blocks 
from the beginning are discarded. Because the lengths of the integers (i.e. the number of bytes used to represent them) may now be different, the data cannot be stored as it is. Instead a \textit{continuation bit} is added in front of each block to form 
a $chunk$. This bit is set to 0 on all chunks except for the last one which contains the least significant bits. The continuation bit is used in decoding to signal whether or not the current integer 
continues in the next chunk.

For example, the standard 16-bit representation of an unsigned integer 42 is \texttt{00000000 00101010}. Assuming the block length is 4, it is split to \texttt{0000 0000 0010 1010}. The empty blocks 
are removed from the beginning and then the continuation bits are added. 1 is added in front of the last block (containing the least significant bits) and 0 to the other blocks, resulting in 
\texttt{\underline{0}0010 \underline{1}1010}. Table~\ref{table:vbytes} contains examples of VB encoded integers with block length 4.

@@

A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length. Since the block 
length is 7, the continuation bit is added in row 8 by adding $2^7$ = 128 to the last block.

@@

Smaller block lengths can yield a better compression rate at the cost of more bit manipulation and therefore possibly slower decompression, while bigger block lengths need less bit manipulation and 
offer less effective compression. On the other hand, a bigger block length means a smaller percentage of added continuation bits.
Generally, a block length of 7 has been popular because it tends to perform well on average and handling chunks as bytes is 
convenient \citep{Man08}.

\section{Variable-byte decoding}
VB decoding reverses the encoding steps: encoded chunks are read until a chunk with 1 as continuation bit is found. Continuation bits 
are removed from all the chunks and the resulting blocks are concatenated to form the original integer. A pseudo code implementation of VB decoding with a block length of 7 is shown in Figure~\ref{vbyte_dec}.
\textsc{Append} adds an element to the end of the list. If the block length is changed, additional bit operation steps when reading the data may be needed.

Continuing from where the encoding example ended, the encoded message was \texttt{00010 11010} with block length of 4 and the goal is to decode a 16 bit unsigned integer. The block from the first chunk is 
extracted and added to $n$, making $n$ = \texttt{10} (bit representation of 2).
The continuation bit was 0 in this chunk, which means the encoded integer continues to the next block. A bitwise shift to the left equal to block length is applied to $n$, 
changing $n$ = \texttt{100000} (bit representation of 32). Then the chunk reading process is repeated. The block of the next chunk is added to $n$, making 
$n$ = \texttt{101010} (bit representation of 42). The continuation bit of this chunk is 1, which means the decoding for this number is complete.

@@

To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
created in addition to the existing data. The length of the bit array in these cases usually has to be close to the length of the data, because encoded element lengths vary. This spatial increment makes 
the compression ineffective and therefore $rank$ and $select$ cannot be applied to any given compression algorithm.

VB encoding has several advantages with $select$ and $rank$: the data is compressed in 
blocks of equal length which significantly shrinks the length of $B$. Moreover the bit array $C$ formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the number of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The rank and select implementations used in this thesis are from C++ library 'SDSL-lite' by \citep{gbmp2014sea}. The library has an implementation of a bit array and several implementations
 of both rank and select to support the bit array. A few useful functions from the library were also used during the implementation phase. Table~\ref{table:supportsize} has $rank$ and $select$ 
size requirements of the implementations used. Both rank implementations have a constant space requirement over the bit array, while select's needed size depends on 
the number of 1's in the data. This chapter describes a way to achieve fast random access to variable length codes. The described method also has a low space overhead, increasing the size of the 
compressed sequence by much less than one bit per element. Two versions of $rank$ from the 'SDSL-lite' library are used in this thesis, $rank_v$ and $rank_{v5}$ and the $select$ version is $select_{MCL}$. 
These implementations were chosen because they are fast and can be created over a regular bit array.

@@

For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, with both division operations being integer divisions. Then 
the number of 1 bits from index $r$ to $i$ are calculated from the original bit array. These three values add up to $rank$($i$). $Rank_{v5}$ is a lighter structure: its superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, 
but reduces memory requirements to a quarter.

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $\log_2 512$ = 9 bits. 
The first relative block always equals 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.

The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $\log_2 2048$ = 11 bits each. The first relative block again equals 0, so the rest 
of the blocks fit to a 64-bit word. The $rank_{v5}$ implementation requires an additional (64+64)/2048 = 6.25\% space.

The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and the location of every 64th set bit is stored 
relative to the superblock. With similar calculations, the location of the closest 64th set bit is calculated and then the bit array is iterated until the required amount of bits is reached. The $select_{MCL}$
 is a practical variant of a PAT tree, the data structure described by \cite{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.

For both rank and select, one function call always gets a value from the superblock and then from the superblock's relative block. The only variable factor is the manual bit count from relative
block's index to wanted index. This is at most the size of one relative block and thus both of the functions can be made work in constant time using modern popcount instructions \citep{Gon05}. The 
implementations are explained in greater detail in \cite{gbmp2014sea}.

@@

The ability to process large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to reduce the size of data so that it fits 
into cache, memory or even hard drive while still allowing it to be accessed easily. Directly addressable codes (DAC), a direct access to any element in a compressed list or array, is 
one of the basic building blocks in compressed data structures. For example, it is needed in inverted index compression \citep{Cul07} and compressed text search \citep{Mou00}.

It is not natively possible to decode the $i$-th element in variable-byte compression algorithms, because the position of the element in the compressed
list depends on the length of the preceding compressed data. Direct access is achievable with supporting data structures. A naive solution would be to store the location 
of each element to an array, but this adds a very large overhead which removes the benefit of compression.

@@

After the data has been arranged this way, a $rank$ data structure is built for each bit array. The data structure does not require additional space on top of that used by the normal VB 
encoding apart from that used by the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

\end{tikzpicture}
\caption{Data structure by \citeauthor{Bri09}, visualized}\label{bris_ds}

@@

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ is the index of desired element's next block in array $A_2$. This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current 
block) is 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.

As an example, the fourth element from the data structure shown in Figure~\ref{bris_ds} is decoded. The first block of data is fetched by directly accessing $A_1$[3]. Then the 
continuation bit $B_1$[3] is checked. It is not set, which indicates that the element continues to the next block. The index of the next block is obtained via a \textsc{Rank}($B_1$, 3) call. There are two zero 
bits in $B_1$ before $B_1$[3], which means the next block is in $A_2$[2]. The continuation bit at $B_2$[1] is 1 so the decompression is completed.

Interestingly, the block length of this method can be allowed to change within different array levels which opens a possibility for further compression optimization. \citeauthor{Bri09} experiments with a few different block 
length setups and leaves open the problem of finding optimal block lengths. While the problem of finding the optimal block length for each data set in regular VB encoding is interesting, the possibility to
optimize it for each block level separately is surely more challenging.
\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}
\State $index \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number + block$
\end{algorithmic}
\end{minipage}
\caption{Example pseudo code of DAC with rank by \citeauthor{Bri09}} \label{bris_pseudo}

@@

\chapter{DAC with select query}
\label{chap:select_impl}
Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the aforementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of just a few bit operations.

@@

Unlike the standard VB encoding, the continuation bits are removed from the chunks and stored in their own bit array, which leaves the blocks in their own array. This allows the 
compressed number to be read from the memory block and removes the need to do block concatenation. The data in blocks is written into memory as they appear in the original integer, so that 
when reading a word from the block byte array, the bits and bytes are already in correct order.

For example, let the integer bit length be 32 and the original data be compressed to 5 blocks of 4 bits. First, the byte location of the first block is needed. $Select_1$($i$) gets the index location
of needed block, but since the block length differs from byte length, the starting byte location $s$ needs to be calculated $s = (select_1(i) \times 4) \div 8$, the latter operand being for integer division.
The first block might not be at the start of the byte $s$. This is why offset $o = (select_1(i) \times 4)$ $mod$ $8$ is needed (in the example case, $o$ is equal to either 0 or 4). Then a 32-bit word $w$ is read from memory from byte location $s$.

At this point, $w$ contains the wanted bits, but also has extra bits from the previous and next compressed integers. If the offset was not zero, there are $o$ trailing bits in $w$ from the previous integer. 
These can be conveniently removed by bit-shifting $w$ right for $o$ bits. Then a pre-calculated bitmask (0xFFFFF in this case) is applied and $w$ contains the desired value. Most systems are
able to read memory only from byte addresses, which leads to a corner case where the compressed integer is stored in more bytes than are contained in a word. In the aforementioned situation, this 
happens if the compressed length was 8 blocks and offset was 4. In this case, another step of reading another byte and storing the remaining bits to $w$ is required. This is entirely avoidable if the integer 
length can be constrained. If the maximum integer length was set to 28 bits, the bits would be within a single word in memory and reading the extra byte would not be needed.

The most intuitive way to calculate the block length of the $i$-th element is to subtract $select_1$(i) from $select_1$(i+1). This however causes an additional 
$select_1$ query, which is costly. A much faster option is to calculate the block length from the continuation bit array. This can be done by reading a word from the bit array and bit-shifting the word 
until the bit representing the start of the $i$-th element is on the rightmost end and then counting the trailing zeros (for example, by using the GCC function \texttt{\_\_builtin\_ctz}). The trailing zeros equal to the number of blocks in the $i$-th element that do not have the continuation bit set. 
This number is then incremented by one to also count the ending block which has the continuation bit set to 1.

Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the pseudo code, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns 
a bit mask for k bytes. In line 8, a long word is read from memory starting at memory address $A$[$begin$].

@@

Figure~\ref{VBarrays} portrays an example of the bit array and the block array. $Select_1(k-1)$+1 returns the index of the starting byte of ith compressed element. The length of the compressed 
element is 3, which is obtainable, e.g., via $select_1(k)$. Then a word is read from the block array, starting from block $A_i$ and blocks not belonging to the element
are removed for example with a bit mask, resulting in k-th element being decoded.

@@

\draw[arrow] (Btarget.south) to (Atarget.north);
\end{tikzpicture}
\caption{Relation of continuation bits and data blocks in VB encoding.} \label{VBarrays}
\end{figure}

\section{Elias-Fano encoding}
Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted. This is usually a given in inverted index lists of search engines. In Elias-Fano encoding \citep{Eli74, Vig13}, 
the data is at first split into low and 
high bits. The low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as a unary bit array B = \texttt{0100100100010001}.
